---
title: "Basic Usage of fashr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
knitr::opts_chunk$set(fig.width = 8, fig.height = 6)
library(fashr)
```

```{r}
# Plot power against FDR
power_versus_fdr <- function(fdr_result, true_indices, fdr_vec = seq(0.01,0.99,by = 0.01), plot = TRUE){
  power <- numeric(length(fdr_vec))
  for (i in 1:length(fdr_vec)){
    significant_indices <- fdr_result$index[fdr_result$FDR < fdr_vec[i]]
    power[i] <- length(intersect(significant_indices, true_indices))/length(true_indices)
  }
  result <- data.frame(fdr = fdr_vec, power = power)
  if (plot){
    plot(fdr_vec, power, type = "l", xlab = "FDR", ylab = "Power")
  }
  return(result)
}

# Plot calibration of FDR
calibration_fdr <- function(fdr_result, true_indices, fdr_vec = seq(0.01,0.99,by = 0.01), plot = TRUE){
  true_discovery_rate <- numeric(length(fdr_vec))
  for (i in 1:length(fdr_vec)){
    significant_indices <- fdr_result$index[fdr_result$FDR < fdr_vec[i]]
    true_discovery_rate[i] <- length(intersect(significant_indices, true_indices))/length(significant_indices)
  }
  result <- data.frame(fdr = fdr_vec, Tfdr = (1 - true_discovery_rate))
  if (plot){
    plot(Tfdr ~ fdr_vec, type = "l", xlab = "nominal FDR", ylab = "observed FDR", data = result)
    lines(c(0,1), c(0,1), col = "red", lty = 2)
  }
  return(result)
}
```

## Setup

We consider the effect size estimate of $100$ eQTLs measured in day $t = 1$ to day $t = 16$:

- there are $50$ eQTLs that are not dynamic, i.e., the effect size is constant over time (Category A).
- there are $30$ eQTLs that are linear dynamic, i.e., the effect size is changing linearly over time (Category B).
- there are $20$ eQTLs that are non-linear dynamic, i.e., the effect size is changing non-linearly over time (Category C).

```{r}
set.seed(1)
N <- 100
propA <- 0.5; propB <- 0.3; propC <- 0.2
sigma_vec <- c(0.05, 0.1, 0.2)

sizeA <- N * propA
data_sim_list_A <- lapply(1:sizeA, function(i) simulate_process(sd_poly = 0.2, type = "nondynamic", sd = sigma_vec, normalize = TRUE))

sizeB <- N * propB
if(sizeB > 0){
data_sim_list_B <- lapply(1:sizeB, function(i) simulate_process(sd_poly = 1, type = "linear", sd = sigma_vec, normalize = TRUE))

}else{
  data_sim_list_B <- list()
}

sizeC <- N * propC
data_sim_list_C <- lapply(1:sizeC, function(i) simulate_process(sd_poly = 0, type = "nonlinear", sd = sigma_vec, sd_fun = 1, p = 1, normalize = TRUE))

datasets <- c(data_sim_list_A, data_sim_list_B, data_sim_list_C)
labels <- c(rep("A", sizeA), rep("B", sizeB), rep("C", sizeC))
indices_A <- 1:sizeA
indices_B <- (sizeA + 1):(sizeA + sizeB)
indices_C <- (sizeA + sizeB + 1):(sizeA + sizeB + sizeC)

par(mfrow = c(3, 3))
for(i in indices_A[1:3]){
  plot(datasets[[i]]$x, datasets[[i]]$y, type = "p", col = "black", lwd = 2, xlab = "Time", ylab = "Effect Size", ylim = c(-1.5, 1.5), main = paste("Category A: ", i))
  lines(datasets[[i]]$x, datasets[[i]]$truef, col = "red", lwd = 1)
}

for(i in indices_B[1:3]){
  plot(datasets[[i]]$x, datasets[[i]]$y, type = "p", col = "black", lwd = 2, xlab = "Time", ylab = "Effect Size", ylim = c(-1.5, 1.5), main = paste("Category B: ", i))
  lines(datasets[[i]]$x, datasets[[i]]$truef, col = "red", lwd = 1)
}

for(i in indices_C[1:3]){
  plot(datasets[[i]]$x, datasets[[i]]$y, type = "p", col = "black", lwd = 2, xlab = "Time", ylab = "Effect Size", ylim = c(-1.5, 1.5), main = paste("Category C: ", i))
  lines(datasets[[i]]$x, datasets[[i]]$truef, col = "red", lwd = 1)
}

par(mfrow = c(1, 1))
```

Let's take a look at the data structure:

```{r}
length(datasets)
str(datasets[[1]])
```

Take a look at the true label of the datasets:

```{r}
table(labels)
```

## Fitting FASH

The default way of fitting FASH is to input the list of datasets
(`data_list`), and specify the column names for the effect size (`Y`),
the standard deviation of the effect size (`S`), and the time points
(`smooth_var`).

The computation could be paralleled by specifying the number of cores
(`num_cores`). Reducing the number of basis functions (`num_basis`)
can also greatly speed up the computation.

### Testing non-linearity

Let's first use the IWP2 model (`order = 2`), and try to detect the
dynamic eQTLs with non-linear dynamics (Category C).

```{r}
fash_fit <- fash(Y = "y", smooth_var = "x", S = "sd", data_list = datasets, 
                  num_cores = 2, num_basis = 20, grid = seq(0, 1, by = 0.025),
                  verbose = TRUE)
fash_fit
```

Let's take a look at the estimated prior:

```{r}
fash_fit$prior_weights
```

We can take a look at their posterior weights in each GP component:

```{r}
plot(fash_fit)
```

The `plot` function automatically plots the posterior weights of the GP components, with darker colors indicating larger PSD (e.g. larger $\sigma_k$). 
The observations are ordered based on posterior mean of the PSD (i.e. $\sigma_k$) in descending order from top to bottom. 
We could also customize the plot in various different ways:

```{r}
plot(fash_fit, discrete = TRUE, ordering = "lfdr")
```

The above structure plot ignores the continuous structure of the PSD (`discrete = TRUE`) and orders the observations based on the local false discovery rate (LFDR) of $p(\beta_i(t)\in S_0|\boldsymbol{Y}, \hat{g})$.
In this case, since the order of IWP is `p = 2`, the base model $S_0$ denotes the space of linear function.


We can then use `fdr_control` to test the null hypothesis that $H_0: \beta_i(t) \in S_0$ at a given FDR level:

```{r}
fdr_result <- fdr_control(fash_fit, alpha = 0.1, plot = TRUE)
```

There are `r sum(fdr_result$fdr_results$FDR < 0.1)` eQTLs flagged as significant at FDR level 0.1. We can take out the indices of these eQTLs:

```{r}
detected_indices <- fdr_result$fdr_results$index[fdr_result$fdr_results$FDR < 0.1]
```

How many of the true (non-linear) dynamic eQTLs are detected?

```{r}
sum(labels[detected_indices] == "C")/sizeC
```

What is the false discovery rate?

```{r}
sum(labels[detected_indices] != "C")/length(detected_indices)
```

Let's take a look at the inferred eQTL effect $\beta_i(t)$ for the detected eQTLs. 

```{r}
fitted_beta <- predict(fash_fit, index = detected_indices[1])
str(fitted_beta)
```

The `predict` function returns the posterior information of the effect size $\beta_i(t)$ for the eQTL specified by `index`.
By default, it returns the posterior of the effect size at each observed time point.
We can also specify the time points to predict the effect size at:

```{r}
fitted_beta_new <- predict(fash_fit, index = detected_indices[1], smooth_var = seq(0, 16, length.out = 100))
str(fitted_beta_new)
```

It is also possible to store M posterior samples rather than the posterior summary:

```{r}
fitted_beta_samples <- predict(fash_fit, index = detected_indices[1], 
                               smooth_var = seq(0, 16, length.out = 100), 
                               only.samples = TRUE, M = 50)
str(fitted_beta_samples)
```

Let's plot the inferred effect size for the first detected eQTL:

```{r}
plot(datasets[[detected_indices[1]]]$x, datasets[[detected_indices[1]]]$y, type = "p", col = "black", lwd = 2, xlab = "Time", ylab = "Effect Size")
lines(fitted_beta_new$x, fitted_beta_new$mean, col = "red", lwd = 2)
lines(datasets[[detected_indices[1]]]$x, datasets[[detected_indices[1]]]$truef, col = "black", lwd = 1, lty = 2)
polygon(c(fitted_beta_new$x, rev(fitted_beta_new$x)), c(fitted_beta_new$lower, rev(fitted_beta_new$upper)), col = rgb(1, 0, 0, 0.2), border = NA)
```

### Testing dynamic eQTLs

What if we are interested in detecting any dynamic eQTLs, not just the
non-linear ones? (i.e., Categories B and C)

We can specify the order of the IWP model to be 1, which will setup a
base model $S_0$ as the space of constant functions.

```{r}
fash_fit_2 <- fash(Y = "y", smooth_var = "x", S = "sd", data_list = datasets, 
                  likelihood = "gaussian", order = 1, pred_step = 1,
                  num_cores = 2, num_basis = 20, grid = seq(0, 1, by = 0.025),
                  verbose = TRUE)
fash_fit_2
```

Take a look at the structure plot:

```{r}
plot(fash_fit_2, discrete = TRUE, ordering = "mean")
```

Let's test the null hypothesis that $H_0: \beta_i(t) \in S_0$ at a
given FDR level:

```{r}
fdr_result_2 <- fdr_control(fash_fit_2, alpha = 0.1, plot = TRUE)
detected_indices_2 <- fdr_result_2$fdr_results$index[fdr_result_2$fdr_results$FDR < 0.1]
```

How many of the true dynamic eQTLs are detected?

```{r}
sum(labels[detected_indices_2] != "A")/(sizeB + sizeC)
```

What is the false discovery rate?

```{r}
sum(labels[detected_indices_2] == "A")/length(detected_indices_2)
```


